---
title: "Project 2"
author: "Ruofei(Sophia) Du, Meghan Phillips, Harivinay Gujjula"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes some helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# You will need to install it (once) from GitHub.
# library(devtools)
# devtools::install_github("physicsland/ezids")
# Then load the package in your R session.
library(ezids)
library(dplyr)
library(ggplot2)
library(rpart)
library(rpart.plot)
library(tidyr)
library(kableExtra)
```

```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=F, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(results="markup", warning = F, message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

# Introduction 
Our dataset was obtained from Kaggle by researching music-related dataset. 
The dataset contains more than 114,000 observations and 21 variables related to Spotify tracks. These variables include track information such as track name and artist name, the popularity of the songs, as well as several audio features, including energy, danceability, valence, tempo, and other characteristics that describe the musical properties of each song.
In addition, the dataset is evenly distributed across genres, which allows for fair comparisons when analyzing the effect of genre on popularity.

```{r, include = F}
spo <- read.csv("spotifydataset.csv")
str(spo)
summary(spo)
dim(spo)
```

# Research Questions

There are many possible ways to explore this dataset. In this project, we are particularly interested in understanding what makes a song popular.

First, we are interested in whether the audio features affects a song's popularity. By analyzing the relationship between popularity and audio features, we learn whether listeners are primarily attracted to the musical characteristics of a track .
Next, we explore whether genre plays a role in song popularity. Since artist popularity may influence the popularity of a track, we exclude artist-related variable in our analysis to focus only on the audio features of the songs themselves. 
Finally, we investigate whether the explicitness has any additional effect on popularity beyond audio features and genre.

This leads to several research questions:

What are the distributional characteristics of key variables (popularity, energy, danceability, and valence), and are any transformations or outlier treatments needed prior to modeling? (EDA) - Vinay and Meg
Do audio features exhibit linear relationships with track popularity? (simple linear regression, scatter plots or something?) - Vinay
How well does each individual audio feature predict popularity using simple linear regression models? (4 audio feature vs popularity) - Vinay
How well does audio feature and genre predict popularity using linear regression models? (audio features+ genre vs polularity) - Sophia
How does genre influence track popularity, and does including genre improve the explanatory power of the model? (comparison and tree) - Sophia
After controlling for genre, do audio features remain significant predictors of popularity? 
Does including genre and explicitness improve the model’s ability to predict popularity?	(audio features+ genre + explicitness vs popularity) - Meg
(Conclusion - Meg)


# Data Cleaning
To investigate these research questions, we first check to see if we need to clean the dataset and remove columns that are not relevant to our analysis.

From the summary, we can see that some tempo have a value of 0. However, the tempo (BPM) of a song cannot realistically be 0. Although this value is not recorded as NA in the dataset, it still represents an invalid observation. At this point, although tempo = 0 is not realistic, we retain these observations initially to avoid unnecessary data loss, and assess their impact during modeling.

Based on the information provided in the dataset, we focus on variables that may influence song popularity. In particular, we examine whether audio features or genre have any effect on the popularity of songs. Therefore, we select the relevant variables, including popularity, several audio features (danceability, energy, valence, and tempo), the explicit content indicator, and the track genre.

Variables such as track name and artist are excluded, as they are not directly used in modeling process. As no missing values were detected in the selected variables, no additional removal of observations was required. 

```{r}
spo1 <- spo[, c(
  "popularity",
  "danceability",
  "energy",
  "valence",
  "tempo",
  "explicit",
  "track_genre"
)]
```
```{r, include=FALSE}
str(spo1)
colSums(is.na(spo1))
```
```{r, include=FALSE}
spo1$explicit <- as.factor(spo1$explicit)
```
```{r}
sum(spo1$tempo == 0)
```
```{r, include=FALSE}
write.csv(spo1,"cleaned_spodata.csv")
```

# EDA

```{r}
ggplot(spo1, aes(x = popularity)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  theme_minimal()

 numeric_vars <- spo1 %>% select(where(is.numeric))

cor_matrix <- cor(numeric_vars, use = "complete.obs")
cor_with_pop <- cor_matrix["popularity", ]
sort(cor_with_pop, decreasing = TRUE)

spo1 %>%
  group_by(track_genre) %>%
  summarise(avg_popularity = mean(popularity, na.rm = TRUE)) %>%
  arrange(desc(avg_popularity)) %>%
  slice_head(n = 25) %>%
  ggplot(aes(x = reorder(track_genre, avg_popularity), y = avg_popularity)) +
  geom_col() +
  coord_flip() +
  theme_minimal()
```
The histogram of track popularity shows a heavily right-skewed distribution, with a large concentration of tracks at very low popularity values (close to 0). A majority of tracks fall within the 20–60 popularity range, while very few tracks achieve extremely high popularity (>80).

Most songs on Spotify are not highly popular, suggesting high-popularity tracks are rare and potentially influenced by external factors


## Linear relationships between each individual audio feature and popularity

```{r}
ggplot(spo1, aes(x = danceability, y = popularity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()

ggplot(spo1, aes(x = energy, y = popularity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()

ggplot(spo1, aes(x = valence, y = popularity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()

ggplot(spo1, aes(x = tempo, y = popularity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()
```
The scatterplots of popularity against danceability, energy, valence, and tempo all show highly dispersed point clouds with no clear linear pattern, indicating weak relationships between these audio features and popularity. While the fitted regression lines (in red) suggest slight trends—such as a small positive association for danceability and tempo, a negligible relationship for energy, and a slight negative trend for valence—the slopes are nearly flat across all plots. This visual evidence aligns with the regression results, confirming that although some relationships may be statistically significant due to the large sample size, they are not practically meaningful.
## audio feature prediction of popularity using simple linear regression models

```{r}
lm_dance <- lm(popularity ~ danceability, data = spo1)
summary(lm_dance)

lm_energy <- lm(popularity ~ energy, data = spo1)
summary(lm_energy)

lm_valence <- lm(popularity ~ valence, data = spo1)
summary(lm_valence)

lm_tempo <- lm(popularity ~ tempo, data = spo1)
summary(lm_tempo)
```
Although some audio features like danceability and valence are statistically significant predictors of popularity, their explanatory power is extremely low. This indicates that popularity is not driven by individual features, but likely by more complex combinations of characteristics and external factors.
## Audio Features vs Popularity
```{r}
m1 <- lm(popularity ~ danceability + energy + valence + tempo, data = spo1)
summary(m1)
anova(m1)
```
```{r}
library(kableExtra)

coef_table <- data.frame(
  Variable = c("Intercept", "Danceability", "Energy", "Valence", "Tempo"),
  Estimate = c(28.44, 9.67, 0.59, -6.80, 0.018),
  `Std. Error` = c(0.375, 0.438, 0.280, 0.297, 0.0023),
  `p-value` = c("<0.001", "<0.001", "0.035", "<0.001", "<0.001")
)

coef_table %>%
  kbl(caption = "Linear Regression Results (Audio Features Only)",
      align = c("l", "c", "c", "c")) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, background = "gray", color = "white") %>%
  column_spec(2, bold = TRUE)
```
```{r}
anova_table <- data.frame(
  Variable = c("Danceability", "Energy", "Valence", "Tempo"),
  `F value` = c(152.82, 1.10, 502.97, 58.54),
  `p-value` = c("<0.001", "0.295", "<0.001", "<0.001")
)

anova_table %>%
  kbl(caption = "ANOVA Results for Audio Features",
      align = c("l", "c", "c")) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, background = "gray", color = "white") %>%
  column_spec(2, bold = TRUE)
```

# Audio Features + Genre vs Popularity
## Linear Regression Model
```{r, include=FALSE}
m2 <- lm(popularity ~ danceability + energy + valence + tempo + track_genre, data = spo1)
summary(m2)
anova(m1, m2)
```
```{r}
genre_highlights <- data.frame(
  Genre = c("pop-film", "k-pop", "sad", "grunge", "indian",
            "—",
            "iranian", "romance", "latin", "detroit-techno", "chicago-house"),
  Coefficient = c("+17.22", "+14.66", "+8.84", "+7.95", "+7.11",
                  "—",
                  "-39.77", "-38.59", "-34.03", "-32.14", "-30.60"),
  Category = c(rep("Top 5", 5), "—", rep("Bottom 5", 5))
)

genre_highlights %>%
  kbl(caption = "Top 5 and Bottom 5 Genres by Coefficient (vs. Acoustic baseline)",
      align = c("l", "c", "c"),
      col.names = c("Genre", "Coefficient", "")) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, background = "grey", color = "white") %>%
  row_spec(1:5, background = "lightblue") %>% 
  row_spec(6, background = "white") %>% 
  row_spec(7:11, background = "lightpink") %>%
  column_spec(2, bold = TRUE)
audio_coef <- data.frame(
  `Audio Feature` = c("Danceability", "Energy", "Valence", "Tempo"),
  Coefficient = c("+7.49", "-1.01", "-5.58", "+0.009"),
  `p-value` = c("< 0.001", "0.002", "< 0.001", "< 0.001"),
  check.names = FALSE
)

audio_coef %>%
  kbl(caption = "Audio Feature Coefficients (Model 2)",
      align = c("l", "c", "c")) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, background = "gray", color = "white") %>%
  column_spec(2, bold = TRUE)
model_comp <- data.frame(
  ` ` = c("R²", "Adj R²", "RSE", "F-statistic"),
  `Model 1 (Audio Only)` = c("0.621%", "0.621%", "22.25", "178.9"),
  `Model 2 (Audio + Genre)` = c("25.74%", "25.66%", "19.24", "336.9"),
  check.names = FALSE
)
model_comp %>%
  kbl(align = c("l", "c", "c")) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, background = "gray", color = "white") %>%
  column_spec(3, background = "lightblue", bold = TRUE)
```

The multiple regression model includes both audio features and genre as predictors for popularity. Comparing the first model with only audio features as predictors, the second model shows a great improvement in explanatory power. The R-squared value increased from 0.00621 to 0.2566. This means that model 2 could explain about 25.66% of the variation in track popularity, while the model 1 could only explain 0.62% of the variance in track popularity. 

The inclusion of genres significantly improves our model, suggesting that genre is a key predictor of song popularity. The coefficient for genre vary widely, which suggests the large differences in popularity across different music genres. Although some of the audio features such as danceability and valence has a small p-value and are statistically significant to predicting the popularity, their effect sizes are much smaller than including genres. Although audio features contributes in predicting the popularity, they are not the primary predictors. 

Overall, the result of model 2 indicates that genres are important predictors compared to audio features. 

## Visualization
### Top 10 Genre and Bottom 10 Genres
```{r}
genre_avg <- spo1 %>%
  group_by(track_genre) %>%
  summarise(mean_pop = mean(popularity)) %>%
  arrange(desc(mean_pop))
top10 <- genre_avg %>%
  slice_head(n = 10)
bot10 <- genre_avg %>%
  slice_tail(n = 10)
```
To better visualize our result, we calculate the average popularity of each genre and arrange them in a descending order. 
```{r}
ggplot(top10,
       aes(x = reorder(track_genre, mean_pop),
           y = mean_pop,
           fill = mean_pop)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Genres Ranked by Average Popularity",
    x = "Genre",
    y = "Average Popularity of the genre"
  ) +
  theme_minimal() +
  scale_fill_gradient(low = "lightblue", high = "darkblue")
```
```{r}
ggplot(bot10,
       aes(x = reorder(track_genre, mean_pop),
           y = mean_pop,
           fill = mean_pop)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Bottom 10 Genres Ranked by Average Popularity",
    x = "Genre",
    y = "Average Popularity of the genre"
  ) +
  theme_minimal() +
  scale_fill_gradient(low = "lightpink", high = "salmon")
```

The bar charts display the top and bottom 10 genres ranked by their average popularity. A clear disparity can be observed between the most and least popular genres. Popular genres such as pop-film and k-pop have an average popularity score close to 60， while on the other side, unpopular genres like iranian and romance have average popularity score below 10. 

The large differences between the top genres and bottom genres suggest that genre is a dominant factor influencing track popularity.

The result strongly align with the result of our regression model findings, which concludes that the inclusion of genre led to a substantial improvement in the model. The wide range of average popularity across different genres suggest that genre is a significant factor influencing the popularity.

The result is consistent with earlier exploratory analysis. 

## Regression Tree
```{r}
set.seed(1123)

tree_model <- rpart(
  popularity ~ danceability + energy + valence + tempo + track_genre + explicit,
  data = spo1,
  method = "anova",
  control = rpart.control(maxdepth = 3)
)

rpart.plot(
  tree_model,
  type = 2,
  extra = 0,
  fallen.leaves = TRUE,
  cex = 1,
  box.palette = "Blues",
  branch.lty = 3,
  shadow.col = "gray"
)
```
```{r, include = F}
path.rpart(tree_model, nodes = rownames(tree_model$frame))
```

A regression tree was grown to further understand how genre and audio features influence the popularity. In our tree here, genre is the primary splitting factor, and we reach a max depth of 3, which resulted in 4 branches in the end. The tree tries to split the genres based on their popularity and minimize the variance in popularity within the leaf. 

The resulting branches show clear differences in average popularity levels. The lowest group has an average popularity of around 12 and is dominated by slower, more traditional genres such as classical and jazz. The lower-mid group has an average of around 25 and includes more niche and moderately upbeat genres. The upper-mid group reaches around 36 and contains even more contemporary genres such as indie and hip-hop. The highest popularity group has an average around 46 and is dominated by mainstream and highly popular genres such as pop, K-pop, and electronic music.

Overall, the tree suggests a clear gradient in popularity across genre groups, with more mainstream and modern genres associated with higher popularity levels.

## Audio Features by Branches (Standardized)
```{r}
spo1$leaf <- tree_model$where
branch_summary <- spo1 %>%
  group_by(leaf) %>%
  summarise(
    danceability = mean(danceability),
    energy = mean(energy),
    valence = mean(valence),
    tempo = mean(tempo),
    .groups = "drop"
  ) %>%
  mutate(
    leaf_label = case_when(
      leaf == 3 ~ "Low",
      leaf == 4 ~ "Lower-Mid",
      leaf == 6 ~ "Upper-Mid",
      leaf == 7 ~ "High"
    )
  )
branch_scaled <- branch_summary %>%
  mutate(
    danceability = scale(danceability)[,1],
    energy = scale(energy)[,1],
    valence = scale(valence)[,1],
    tempo = scale(tempo)[,1]
  )
branch_long <- branch_scaled %>%
  pivot_longer(
    cols = c(danceability, energy, valence, tempo),
    names_to = "feature",
    values_to = "value"
  )
branch_long$leaf_label <- factor(
  branch_long$leaf_label,
  levels = c("Low", "Lower-Mid", "Upper-Mid", "High")
)
ggplot(branch_long, aes(x = feature, y = leaf_label, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  labs(
    title = "Audio Features by Tree Branch (Standardized)",
    x = "Audio Feature",
    y = "Tree Branch"
  ) +
  theme_minimal()
```

To better understand the characteristics of each branch, we examine the average audio features within each leaf using a standardized heatmap. The heatmap showss clear differences in audio characteristics across the four branches identified by the regression tree. 

The lowest popularity group is associated with consistently low levels of energy and tempo, suggesting that less dynamic tracks tend to perform poorly. In contrast, the lower-mid and upper-mid groups show higher values across most audio features, suggesting that more engaging and rhythmically active songs are more likely to achieve moderate levels of popularity. 

However, it is particularly interesting to observe that the highest popularity group has a higher tempo but relatively lower valence, indicating that fast-paced songs do not necessarily need to be emotionally positive to be successful. 

This suggests that highly popular songs do not share a single, consistent audio profile. Instead, different combinations of audio features may lead to high popularity. This variability helps explain why audio features alone account for only a small proportion of the variation in popularity in our regression models.

Overall, these patterns suggest that while genre determines the primary grouping of tracks, audio features help explain differences in popularity within each group in some way. 

## Logistic Regression
```{r, include=FALSE}
spo1$pop_any <- ifelse(spo1$popularity > 0, 1, 0)
spo1$pop_high <- ifelse(spo1$popularity >= 60, 1, 0)
model_logit_any <- glm(
  pop_any ~ danceability + energy + valence + tempo + track_genre + explicit,
  data = spo1,
  family = "binomial"
)
model_logit_high <- glm(
  pop_high ~ danceability + energy + valence + tempo + track_genre + explicit,
  data = spo1,
  family = "binomial"
)
summary(model_logit_any)
summary(model_logit_high)
```
```{r}
library(kableExtra)

logit_table <- data.frame(
  Feature = c("Danceability", "Energy", "Valence", "Tempo", "Explicit (True)"),
  `Any Popular (>1)` = c(1.66, 1.07, 0.53, 1.002, 1.21),
  `High Popular (≥60)` = c(2.03, 1.02, 0.78, 1.00, 1.38)
)

logit_table %>%
  kbl(
    caption = "Logistic Regression: Odds Ratios for Audio Features",
    align = c("l", "c", "c")
  ) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, background = "gray", color = "white") %>%
  column_spec(2:3, bold = TRUE) %>%
  footnote(
    general = "Odds ratios shown. Values >1 indicate positive effects. Baseline: acoustic genre, non-explicit tracks."
  )
```

Based on our exploratory data analysis, we observe that the distribution of popularity is highly skewed, with a large proportion of tracks having zero popularity. This raises an important question regarding how “popularity” should be defined in this context.

To address this, we consider two alternative definitions: whether a track has any popularity (popularity > 0) and whether a track achieves high popularity (popularity >= 60). This distinction is motivated by our earlier tree and heatmap analyses, which suggest that gaining some popularity may involve different mechanisms than achieving very high popularity.

We therefore employ logistic regression models to examine how audio features and genre influence these two outcomes. Logistic regression allows us to model the probability of a track being classified as “popular” under each definition, providing a complementary perspective to our earlier linear and tree-based models. We use this logistic regression model as a tool to better understand the reason why a song is popular, instead of a predicting model, therefore, we did not split the data into train and test sample, but instead using the regression model to compare the effect of predictors on the dependent variable (popularity).

From the results summarized in the table, we find that danceability shows a consistently strong positive effect across both models, indicating that more rhythmically engaging tracks are more likely to gain attention and achieve high popularity. In contrast, valence has a negative effect, suggesting that more emotionally positive songs are not necessarily more successful.

Interestingly, tempo is significant in predicting whether a track gains any popularity, but not in predicting highly popular tracks. This suggests that tempo may help a song attract initial attention, but does not determine whether it becomes widely successful.

Overall, these findings indicate that gaining some popularity and achieving high popularity are driven by different factors. The reason for a song to gain some popularity and become very popular is not quite the same, however, genre remains the dominant influence across both definitions.

# Audio Features + Genre + Explict - Meg 
Does just explicitness help predict popularity?	We know there is statistical correlation, but...
```{r}
exp_lm<- lm(popularity ~ explicit, data=spo1)
summary(exp_lm)
```
outcome: explicitTRUE coefficient is 3.516, meaning on average explicit songs are about 3.5 points higher in popularity than non-explicit. The p-value is incredibly small, meaning there is statistical significance. However, the R-squared model fit is showing that only 19% of the variation in popularity is explained by whether a song is explicit. Also, the Residual Standard Error being 22.3 shows high variation. Explicitness alone is not a main driver of if a song has higher popularity.



#Does just genre and explicitness help predict popularity?	
```{r}

gen_exp_lm<- lm(popularity ~ track_genre + explicit, data=spo1)
summary(gen_exp_lm)

```
The adjusted R-squared is 0.255, meaning 25.5% of the occurrences can be explained by 

#ANOVA - is adding explicitness better?
```{r}
anova(exp_lm, gen_exp_lm)
```
F statistic being 343 shows significant increase from the 222 of just explicitness. While explicitness alone explains very little variation in popularity (R² ≈ 0.002), adding track_genre significantly improves the model (F = 343, p < 2e-16) and increases the explained variance by about 25%. This indicates that genre is a strong and meaningful predictor of song popularity.

```{r}
m3 <- lm(popularity ~ danceability + energy + valence + tempo + track_genre + explicit, data = spo1)
summary(m3)
```
```{r}
anova(m1, m2, m3)
```

This shows the model 3 has slight improvement over model 2, but overall, adding the genre coefficient has the greatest impact on the RSS. 

