---
title: "Project 2"
author: "Ruofei(Sophia) Du, Meghan Phillips, Harivinay Gujjula"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes some helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# You will need to install it (once) from GitHub.
# library(devtools)
# devtools::install_github("physicsland/ezids")
# Then load the package in your R session.
library(ezids)
library(dplyr)
library(ggplot2)
library(rpart)
library(rpart.plot)
```

```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=F, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(results="markup", warning = F, message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

# Introduction 
Our dataset was obtained from Kaggle by researching music-related dataset. 
The dataset contains more than 114,000 observations and 21 variables related to Spotify tracks. These variables include track information such as track name and artist name, the popularity of the songs, as well as several audio features, including energy, danceability, valence, tempo, and other characteristics that describe the musical properties of each song.
In addition, the dataset is evenly distributed across genres, which allows for fair comparisons when analyzing the effect of genre on popularity.

```{r, include = F}
spo <- read.csv("spotifydataset.csv")
str(spo)
summary(spo)
dim(spo)
```

# Research Questions

There are many possible ways to explore this dataset. In this project, we are particularly interested in understanding what makes a song popular.

First, we are interested in whether the audio features affects a song's popularity. By analyzing the relationship between popularity and audio features, we learn whether listeners are primarily attracted to the musical characteristics of a track .
Next, we explore whether genre plays a role in song popularity. Since artist popularity may influence the popularity of a track, we exclude artist-related variable in our analysis to focus only on the audio features of the songs themselves. 
Finally, we investigate whether the explicitness has any additional effect on popularity beyond audio features and genre.

This leads to several research questions:

What are the distributional characteristics of key variables (popularity, energy, danceability, and valence), and are any transformations or outlier treatments needed prior to modeling? (EDA) - Vinay and Meg
Do audio features exhibit linear relationships with track popularity? (simple linear regression, scatter plots or something?) - Vinay
How well does each individual audio feature predict popularity using simple linear regression models? (4 audio feature vs popularity) - Vinay
How well does audio feature and genre predict popularity using linear regression models? (audio features+ genre vs polularity) - Sophia
How does genre influence track popularity, and does including genre improve the explanatory power of the model? (comparison and tree) - Sophia
After controlling for genre, do audio features remain significant predictors of popularity? 
Does including genre and explicitness improve the model’s ability to predict popularity?	(audio features+ genre + explicitness vs popularity) - Meg
(Conclusion - Meg)


# Data Cleaning
To investigate these research questions, we first check to see if we need to clean the dataset and remove columns that are not relevant to our analysis.

From the summary, we can see that some tempo have a value of 0. However, the tempo (BPM) of a song cannot realistically be 0. Although this value is not recorded as NA in the dataset, it still represents an invalid observation. At this point, although tempo = 0 is not realistic, we retain these observations initially to avoid unnecessary data loss, and assess their impact during modeling.

Based on the information provided in the dataset, we focus on variables that may influence song popularity. In particular, we examine whether audio features or genre have any effect on the popularity of songs. Therefore, we select the relevant variables, including popularity, several audio features (danceability, energy, valence, and tempo), the explicit content indicator, and the track genre.

Variables such as track name and artist are excluded, as they are not directly used in modeling process. As no missing values were detected in the selected variables, no additional removal of observations was required. 

```{r}
spo1 <- spo[, c(
  "popularity",
  "danceability",
  "energy",
  "valence",
  "tempo",
  "explicit",
  "track_genre"
)]
```
```{r, include=FALSE}
str(spo1)
colSums(is.na(spo1))
```
```{r, include=FALSE}
spo1$explicit <- as.factor(spo1$explicit)
```
```{r}
sum(spo1$tempo == 0)
```
```{r, include=FALSE}
write.csv(spo1,"cleaned_spodata.csv")
```

#EDA


# Audio Features vs Popularity
```{r}
m1 <- lm(popularity ~ danceability + energy + valence + tempo, data = spo1)
summary(m1)
plot(m1)
```

# Audio Features + Genre vs Popularity
## Linear Regression Model
```{r, include=FALSE}
m2 <- lm(popularity ~ danceability + energy + valence + tempo + track_genre, data = spo1)
summary(m2)
```
The multiple regression model includes both audio features and genre as predictors for popularity. Comparing the first model with only audio features as predictors, the second model shows a great improvement in explanatory power. The R-squared value increased from 0.00621 to 0.2566. This means that model 2 could explain about 25.66% of the variation in track popularity, while the model 1 could only explain 0.62% of the variance in track popularity. 

The inclusion of genres significantly improves our model, suggesting that genre is a key predictor of song popularity. The coefficient for genre vary widely, which suggests the large differences in popularity across different music genres. Although some of the audio features such as danceability and valence has a small p-value and are statistically significant to predicting the popularity, their effect sizes are much smaller than including genres. Although audio features contributes in predicting the popularity, they are not the primary predictors. 

Overall, the result of model 2 indicates that genres are important predictors compared to audio features. 

## Visualization
### Top 10 Genre
```{r}
genre_avg <- spo1 %>%
  group_by(track_genre) %>%
  summarise(mean_pop = mean(popularity)) %>%
  arrange(desc(mean_pop))
top10 <- genre_avg %>%
  slice_head(n = 10)
bot10 <- genre_avg %>%
  slice_tail(n = 10)
```
To better visualize our result, we calculate the average popularity of each genre and arrange them in a descending order. 
```{r}
ggplot(top10,
       aes(x = reorder(track_genre, mean_pop),
           y = mean_pop,
           fill = mean_pop)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Genres Ranked by Average Popularity",
    x = "Genre",
    y = "Average Popularity of the genre"
  ) +
  theme_minimal() +
  scale_fill_gradient(low = "lightblue", high = "darkblue")
```
```{r}
ggplot(bot10,
       aes(x = reorder(track_genre, mean_pop),
           y = mean_pop,
           fill = mean_pop)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Bottom 10 Genres Ranked by Average Popularity",
    x = "Genre",
    y = "Average Popularity of the genre"
  ) +
  theme_minimal() +
  scale_fill_gradient(low = "lightpink", high = "salmon")
```
The bar charts display the top and bottom 10 genres ranked by their average popularity. A clear disparity can be observed between the most and least popular genres. Popular genres such as pop-film and k-pop have an average popularity score close to 60， while on the other side, unpopular genres like iranian and romance have average popularity score below 10. 

The large differences between the top genres and bottom genres suggest that genre is a dominant factor influencing track popularity.

The result strongly align with the result of our regression model findings, which concludes that the inclusion of genre led to a substantial improvement in the model. The wide range of average popularity across different genres suggest that genre is a significant factor influencing the popularity.

The result is consistent with earlier exploratory analysis. 

##Linear relationships between each individual audio feature and popularity

```{r}
ggplot(spo1, aes(x = danceability, y = popularity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()

ggplot(spo1, aes(x = energy, y = popularity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()

ggplot(spo1, aes(x = valence, y = popularity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()

ggplot(spo1, aes(x = tempo, y = popularity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()
```
## audio feature prediction of popularity using simple linear regression models

```{r}
lm_dance <- lm(popularity ~ danceability, data = spo1)
summary(lm_dance)

lm_energy <- lm(popularity ~ energy, data = spo1)
summary(lm_energy)

lm_valence <- lm(popularity ~ valence, data = spo1)
summary(lm_valence)

lm_tempo <- lm(popularity ~ tempo, data = spo1)
summary(lm_tempo)
```




## Regression Tree
```{r}
set.seed(1123)

tree_model <- rpart(
  popularity ~ danceability + energy + valence + tempo + track_genre + explicit,
  data = spo1,
  method = "anova",
  control = rpart.control(maxdepth = 3)
)

rpart.plot(
  tree_model,
  type = 2,
  extra = 1,
  fallen.leaves = TRUE,
  cex = 0.6,
  faclen = 0
)
```
```{r}
path.rpart(tree_model, nodes = rownames(tree_model$frame))
```

# Audio Features + Genre + Explict - Meg 
Does just explicitness help predict popularity?	We know there is statistical correlation, but...
```{r}
exp_lm<- lm(popularity ~ explicit, data=spo1)
summary(exp_lm)
```
outcome: explicitTRUE coefficient is 3.516, meaning on average explicit songs are about 3.5 points higher in popularity than non-explicit. The p-value is incredibly small, meaning there is statistical significance. However, the R-squared model fit is showing that only 19% of the variation in popularity is explained by whether a song is explicit. Also, the Residual Standard Error being 22.3 shows high variation. Explicitness alone is not a main driver of if a song has higher popularity.



#Does just genre and explicitness help predict popularity?	
```{r}

gen_exp_lm<- lm(popularity ~ track_genre + explicit, data=spo1)
summary(gen_exp)

```
The adjusted R-squared is 0.255, meaning 25.5% of the occurrences can be explained by 

#ANOVA - is adding explicitness better?
```{r}
anova(exp_lm, gen_exp_lm)
```
F statistic being 343 shows significant increase from the 222 of just explicitness. While explicitness alone explains very little variation in popularity (R² ≈ 0.002), adding track_genre significantly improves the model (F = 343, p < 2e-16) and increases the explained variance by about 25%. This indicates that genre is a strong and meaningful predictor of song popularity.



